---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

This assignment was divided into 5 sections:

1. Loading and preprocessiong the data;
2. Question 1: What is mean total number of steps taken per day?;
3. Question 2: What is the average daily activity pattern?;
4. Imputing missing values;
5. Question 3: Are there differences in activity patterns between weekdays and weekends?.


## Loading and preprocessing the data
Suppose that you already fetched the repository to the local machine and stored all files into "Documents/RepoData"

```{r loadingdata, echo = TRUE, warning=FALSE}
# Loading libraries
library(ggplot2)
library(dplyr)

# Force results to be in English
Sys.setlocale("LC_ALL","English")

# Unzipping the compressed file
unzip("activity.zip", list = FALSE, overwrite = TRUE)

# Loading the .csv file to the raw dataset
raw_dataset <- read.csv("activity.csv")
```

The first 5 lines of `raw_dataset`:

```{r head_raw_dataset, echo = TRUE}
head(raw_dataset,5)
```

Summary of each variable:

```{r summary_raw_dataset, echo = TRUE}
summary(raw_dataset)
```

Type of variable:

```{r type_raw_dataset, echo = TRUE}
# Using the class() function to get the type of each variable
steps_type <- class(raw_dataset$steps)
date_type <- class(raw_dataset$date)
interval_type <- class(raw_dataset$interval)
```

* steps: `r steps_type`
* date: `r date_type`
* interval: `r interval_type`

Fixing the `date` variable to became `as.Date`.

```{r date_correction, echo = TRUE}
raw_dataset$date <- as.Date(raw_dataset$date)
date_type <- class(raw_dataset$date)
```

Now the date variable `date` is `r date_type` type.

Eliminating the `NA` values:
```{r complete_case, echo = TRUE}
# Selection Vector used to eliminate rows with NA steps.
selection_rows <- complete.cases(raw_dataset)

# Cleaned dataset
dataset <- raw_dataset[selection_rows,TRUE]
```

The first 5 lines of the cleaned dataset:
```{r dataset, echo = TRUE}
head(dataset,5)
```

## What is mean total number of steps taken per day?

1. The `dataset` has lines that correspond to a sample of a 5 minutes of the day. It is necessary to `aggregate` the rows with the same day to calculate the average of steps.

```{r aggregate_steps, echo = TRUE}
# The aggregate function was used to sum steps of the same day
dataset_aggregate <- aggregate(steps ~ date, data = dataset, FUN = sum)

# The first 5 lines of the aggregate dataset (total number of steps taken per day)
head(dataset_aggregate,5)
```

The barplot below shows the evolution of `steps` between `r min(dataset_aggregate$date)` and `r max(dataset_aggregate$date)`. This graphic intend to show the `step` variation during the period day by day.

```{r graphic_aggregate_steps, echo = TRUE}
# temporal historic of step using barplot
barplot(dataset_aggregate$steps, names.arg = dataset_aggregate$date, xlab = "date", ylab = "steps",main =  "Temporal serie of steps")

# Mean
average_steps <- mean(dataset_aggregate$steps)

# Horizontal red line to show the step average
abline(h = average_steps,col="red",lwd=4)
```

2. The the `steps` histogram is show below. This histogram intend to show the frequency of daily `steps`, what is the most comom and what is exceptions.

```{r histogram_aggregate_steps, echo = TRUE}
# Plot the histogram
hist(dataset_aggregate$steps, main = "Histogram of steps",xlab = 'Daily Steps', ylab = 'Frequency',breaks = 10)

# Vertical line to show the step average
abline(v = average_steps,col="red",lwd=4)
```

The `summary` show the mean and median of `steps`.
```{r summary_step, echo = TRUE}
# Summary of steps
summary(dataset_aggregate$steps)

# Results
mean_step <- summary(dataset_aggregate$steps)['Mean']
median_step <- summary(dataset_aggregate$steps)['Median']

```

3. Results:
* Mean = `r format(mean_step,digits = 1)`
* Median = `r format(median_step,digits = 1)`

## What is the average daily activity pattern?

Analog of the above implementation, instead of day this question ask to `aggregate` by interval.

```{r aggregate_interval, echo = TRUE}
# The aggregate function was used to "mean" steps of the same interval
dataset_aggregate_interval <- aggregate(steps ~ interval, data = dataset, FUN = mean)

# The first 5 lines of the interval aggregate dataset
head(dataset_aggregate_interval,5)
```

The Time serie of Average Steps by Intreval

```{r time_serie_interval, echo = TRUE}
# Maximum Average by Interval
max_step_average_interval <- max(dataset_aggregate_interval$steps)

# When the max occurs
interval_max_average <- dataset_aggregate_interval[dataset_aggregate_interval$steps == max_step_average_interval,TRUE]$interval

# Line graphic to show the time serie of average step by interval and the red vertical line to mark the maximum average step
ggplot(dataset_aggregate_interval,aes(interval,steps))+
        ggtitle("Time Series of Average Steps by Interval") +
        geom_line() +
        geom_vline(xintercept = interval_max_average , colour = "red" , size = 1)
```

The maximum average step by interval is `r max_step_average_interval` and it is occurs during the interval `r interval_max_average`

## Imputing missing values

Summary the `raw_data`:
```{r missing_values, echo = TRUE}
summary(raw_dataset)
```

Number of `NA` values in `steps`:
```{r NA_step, echo = TRUE}
NA_steps <- summary(raw_dataset$steps)["NA's"]
```
* There are `r NA_steps` intervals with `NA` value.

The strategy to fill the `NA` is based on steps average for each weekday.
```{r strategy, echo = TRUE}
# Using the dplyr package
dataset_aggregate <- tbl_df(dataset_aggregate)

# Adding a new column to keep the weekday
dataset_aggregate <- mutate(dataset_aggregate,weekday = weekdays(dataset_aggregate$date))

# Saving weekdays
weekdays_dataset <- unique(dataset_aggregate$weekday)

# One average for each weekday 
mean_weekday <- vector()
for (i in 1:length(weekdays_dataset) ) {
        filter_dataset_aggregate <- filter(dataset_aggregate,weekday == weekdays_dataset[i])$steps
        
        mean_weekday[i] <- mean(filter_dataset_aggregate)
        
}

names(mean_weekday) <- c(weekdays_dataset)
```

Each weekday mean that will be used to fill `NA` values:

```{r results_weekday_mean, echo = TRUE}
# Mean by interval
mean_weekday <- mean_weekday/288

mean_weekday
```

Substituing the `NA` with the `mean_weekday` values.

```{r NA_to_weekdays, echo = TRUE}

# Transforming the regular dataframe into a dplyr table
raw_dataset_NA_replace <- tbl_df(raw_dataset)

# Mutate the raw_dataset to insert a new column of weekday
raw_dataset_NA_replace <- mutate(raw_dataset_NA_replace,weekday = weekdays(raw_dataset_NA_replace$date))

# Loop to check each row
for (j in weekdays_dataset) {
        raw_dataset_NA_replace <- mutate(raw_dataset_NA_replace,
                              steps = ifelse(test = ( (weekday == j) & (is.na(steps) == TRUE) ),
                                             yes = mean_weekday[[j]] ,
                                             no = steps ) )
        }

```

Comparison with cleaned dataset and dataset with `NA` values replace by strategy adopth.

```{r hist_replace_steps_NA, echo = TRUE}
# The aggregate function was used to sum steps of the same day
dataset_aggregate_NA_replace <- aggregate(steps ~ date, data = raw_dataset_NA_replace, FUN = sum)

# Plot the histogram with NA values replaced regarding the strategy
hist(dataset_aggregate_NA_replace$steps, main = "Histogram of steps - With NA Fill Strategy",
     xlab = 'Daily Steps', ylab = 'Frequency',breaks = 10)

# Red Vertical line to show the step average with NA values replaced regarding the strategy
abline(v = mean(dataset_aggregate_NA_replace$steps),col="red",lwd=4)

# Blue Vertical line to show the step median with NA values replaced regarding the strategy
abline(v = median(dataset_aggregate_NA_replace$steps),col="blue",lwd=4)

```

The `summary` show the mean and median of `steps` after fill all `NA` values by the strategy adopt.
```{r median_mean_after_replace_NA, echo = TRUE}
summary(dataset_aggregate_NA_replace$steps)
```

Comparison before and after fill `NA` values.
```{r comparison_table, echo = TRUE}
# Comparison Table
rbind('Weekday Strategy' = summary(dataset_aggregate_NA_replace$steps),
      'Without NA' = summary(dataset_aggregate$steps))[,3:4]
```

The strategy adopted to fill `NA` values makes the median and mean became farthest, as shown in the table above.

## Are there differences in activity patterns between weekdays and weekends?

Adopting the same srategy to fill NA values.
```{r new_fator, echo = TRUE}
# Removing the additional column created to specify the day of the week in the later question.
raw_dataset_new_factor <- select(raw_dataset_NA_replace,-weekday)

# Mutate the table base on weekday and weekend using the ifelse() function
raw_dataset_new_factor <- mutate(raw_dataset_new_factor,
                                New_factor = ifelse(test = weekdays(raw_dataset_new_factor$date) == c('Saturday','Sunday'), 
                                                    yes = 'weekend', 
                                                    no = 'weekday'))
```

Make a panel plot containing a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r weekday_weekend, echo = TRUE}

raw_dataset_new_factor_aggregate <- aggregate(steps ~ interval + New_factor, data = raw_dataset_new_factor, FUN = mean)


# Line graphic to show the time serie of average step by interval and the red vertical line to mark the maximum average step
g1 <- ggplot(raw_dataset_new_factor_aggregate,aes(interval,steps))+
        ggtitle("Time Series Painel of Average Steps by Interval and Weekdays") +
        geom_line()+
        facet_grid(New_factor ~ .)

print(g1)

```

The people seem to be more active on weekends and we cannot infer any standard activities behavior because all activities was spread on the day. Furthermore, the people seem to be more activities before the work on the weekdays.